Boosted Regression Trees for Data-Driven Auto-Tuning

### 2011-12-11 paper/draft1.tex
paper

@@

This work shows that online auto-tuning can be accelerated by orders of
magnitude by using a regression model built offline as a surrogate for actual
computations on the real hardware.
We use a powerful non-linear regression model (boosted regression trees)
to approximate the empirical relationship between three sets of variables ($\mathcal{A},
\mathcal{B}, \mathcal{C}$) and the observed speed-up over a
reference implementation:
\begin{description}
\item[$\mathcal{A}$:] task description (e.g. shapes, physical layout)
\item[$\mathcal{B}$:] algorithm description (i.e. auto-tuning parameters)
\item[$\mathcal{C}$:] platform description (e.g. micro-benchmarks)
\end{description}
Ideally, \texttt{gpu\_fbcorr} would apply the best
available implementation

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_empirical}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} MeasureTime(a, b, c)$
\algoremark{\color{darkred}{slow}}
\Aitem \Return $b^*$
\end{algorithm}
}
\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_predictive}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $f \setto LoadTimingModel()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
\algoremark{\color{darkred}{fast}}
\Aitem \Return $b^*$
\end{algorithm}
\caption{Pseudo-code template for an auto-tuning implementation of
$\texttt{gpu\_fbcorr}$. Empirical auto-tuning (above) is inevitably slow
because dynamically-generated code must be compiled and run on a number of actual-size inputs.
Predictive auto-tuning (below) can be orders of magnitude faster. We show that it can
also be accurate.
}
\label{fig:argmin}
\end{figure}

>>>>

Current languages and software engineering practices have conspired to make it difficult
to provide high-performance libraries for general mathematical operations on GPU devices.
The dominant approach to implementing high-performance GPU codes has been auto-tuning,
which is fundamentally slow. Compilation with auto-tunin requires running a large number of
sub-compilations and representative computations to actually {\em measure} which
of several possible implementations is fastest. General-purpose libraries such
as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
intimate knowledge of the mathematical problem being solved as well as the
kernel implementations to heuristically choose a good computation strategy quickly when a
user asks for e.g. a matrix product of particular arguments.
Designing and testing libraries of this form is time-consuming, difficult, and provides little
reusable infrastructure to assist in the implementation of different computations.
Libraries for stencil computations and sparse matrix operations must start from scratch.

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_empirical}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} MeasureTime(a, b, c)$
\algoremark{\color{darkred}{slow}}
\Aitem \Return $b^*$
\end{algorithm}
}
\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_predictive}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $f \setto LoadTimingModel()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
\algoremark{\color{darkred}{fast}}
\Aitem \Return $b^*$
\end{algorithm}
\caption{Pseudo-code template for an auto-tuning implementation of
$\texttt{gpu\_fbcorr}$. Empirical auto-tuning (above) is inevitably slow
because dynamically-generated code must be compiled and run on a number of actual-size inputs.
Predictive auto-tuning (below) can be orders of magnitude faster. We show that it can
also be accurate.
}
\label{fig:general_autotune}
\end{figure}

This work shows that auto-tuning can be accelerated by orders of
magnitude by using a regression model built offline as a surrogate for actual
computations on the real hardware.
The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
An auto-tuning based routine must operate on three sets of variables:
\begin{description}
\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
\item[$\mathcal{C}$:] platform description (compute capability, micro-benchmarks)
\end{description}
The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune} 
might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
so it would be all but unusable as a normal library routine.
However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~{\ref:fig:general_autotune}, bottom).

Our contribution of the present work is to show that boosted regression tree models,
a powerful machine learning technique for function approximation, can meet the
requirements of this application.
We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning. XXX numbers:
how good, how fast can prediction be done how much training data is necessary?

\section{Filterbank Correlation}
\label{sec:fbcorr}

@@
